---
title: "Relationships between variables"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```


## Comparing between two numeric variables

### Example analysis

**Here's an example analysis ... **

-1 *Ask your question*

-2 *Check and explore your data*

```{r ,exercise = TRUE}

```

```{r ,exercise = TRUE}

```

```{r -solution, exercise.reveal_solution = TRUE}

```

```{r , exercise=TRUE}

```

```{r , exercise=TRUE}

```

```{r -solution, exercise.reveal_solution = TRUE}

```

-3 *Write your Null Hypothesis - H0*

**There is no association between the woodlouse species and soil type.**
In other words, woodlouse species is independent of soil type.

-4 *Decide on your stats test*

```{r , exercise=TRUE}

```

-5 *Do test and check significance value*

```{r , exercise=TRUE}
```


-6 *Present your results*

Let's come back to a suitable plot for a Chi-square test. Two options are mosaicplots or Cohen plots.

Mosaic plot

```{r mosaic_plot, exercise = TRUE}



```

Cohen Plot

```{r, cohen_plot, exercise = TRUE}

assocplot(as.matrix(woodlice))

```




```{r quiz1}
quiz(
  question_numeric("",
                   answer(, correct = TRUE)
  ),
  question_radio("",
                 answer(),
                 answer(, correct = TRUE),
                 answer(),
                 answer(),
                 answer(),
                 random_answer_order = TRUE,
                 allow_retry = TRUE),
  question_checkbox("Select all ..",
                    answer(, correct = TRUE),
                    answer(, correct = FALSE),
                    answer(, correct = TRUE),
                    random_answer_order = TRUE,
                    allow_retry = TRUE)
  
)
```




## Comparing between two categorical variables

### Example analysis

**Here's an example analysis ... **

-1 *Ask your question*

-2 *Check and explore your data*

```{r ,exercise = TRUE}

```

```{r ,exercise = TRUE}

```

```{r -solution, exercise.reveal_solution = TRUE}

```

```{r , exercise=TRUE}

```

```{r , exercise=TRUE}

```

```{r -solution, exercise.reveal_solution = TRUE}

```

-3 *Write your Null Hypothesis - H0*

There is no difference ...

-4 *Decide on your stats test*

```{r , exercise=TRUE}

```

-5 *Do test and check significance value*

```{r , exercise=TRUE}
```


-6 *Present your results*

```{r quiz2}
quiz(
  question_numeric("",
                   answer(, correct = TRUE)
  ),
  question_radio("",
                 answer(),
                 answer(, correct = TRUE),
                 answer(),
                 answer(),
                 answer(),
                 random_answer_order = TRUE,
                 allow_retry = TRUE),
  question_checkbox("Select all ..",
                    answer(, correct = TRUE),
                    answer(, correct = FALSE),
                    answer(, correct = TRUE),
                    random_answer_order = TRUE,
                    allow_retry = TRUE)
  
)
```

