---
title: "GIS2 Making maps in R"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)

library(sf)
library(terra)
library(tmap)

```

## Making maps in R

#### **Intro the tmap package**

The tmap package can be used to make static maps (e.g. to export to an image for a report or publication), interactive maps (for a webpage or online document) as well as making it easy to make multiple maps at the same time, grouped by a column in your data frame (e.g. by species, by region, by project). 


tmap builds up map in layers (similar to ggplot2, if you are familiar with that package). Each line of code (or pairs, more exactly) adds another layer to your map (similar to turning on layers in ArcGIS).

tmap uses two main types of functions and we often use these together in subsequent lines of code. First, **a function to add spatial data**, and then **a function to change the style of the spatial data** (e.g. choose symbol type, change size of symbols, colour polygons by an attribute, change line thickness, etc).

The first function is generally `tm_shape()` which adds an sf or raster object to your map. The second function depends on the type of spatial data you are adding, some common examples are:


function| geometry type | example functionality
--------|---------------|-----------------------
tm_symbols | points | change point colour, size
tm_dots | points | simple dots
tm_borders | polygons| simple polygons without fill colour
tm_polygons | polygons | complete control of over polygon fill, border, etc
tm_line | lines | colour, thickness and style of lines
tm_raster | raster | raster attributes

##### **Simple example**

```{r tmap_simple, exercise = TRUE}



```



## Make a map in R

Create a map for a report that shows the sample design for either the first or second sampling design question above. Write the code in a new R script and export the map to an image, ready to paste into a report.



## Predicates

We use `st_intersects()` to identify which polygons, points or lines intersect with the bounds of another polygon. Note that this function basically gives us a **"yes / no"** answer, a shape either intersects with another one or not. Conversely, the `st_intersection()` function returns a geometry (e.g. a polygon). We're going to use the circle3 from our `crcls` object to identify (and then extract) the points that lie inside it. First we will select just circle3. Note how we use the `[]`, with a comma, to select just circle3. This codes says *return the row(s) where `id` is equal to "circle3"*.

```{r clip1, exercise = TRUE}

circle3 <- crcls[crcls$id == "circle3",]

```
Have a look at these random points, with circle3 plotted on top:

```{r clip2}
circle3 <- crcls[crcls$id == "circle3",]

plot(st_geometry(rnd_pts), pch = 16, ylim = c(0,15),xlim = c(0,15))
plot(st_geometry(circle3), add= TRUE)
```

In the `st_intersects()` function, the first argument is the shape (or geometry) that we are asking that "yes / no" question about. For example, do the points (1^st^ argument) lie inside the circle (2^nd^ argument)? Run the code and then print the result (i.e. run the `pts_intersect` object).

```{r clip3, exercise = TRUE, exercise.setup = "clip1"}
pts_intersect <- st_intersects(rnd_pts, circle3)

```

```{r clip3-solution, exercise.reveal_solution = TRUE}
pts_intersect <- st_intersects(rnd_pts, circle3)
pts_intersect

```

What can you see?  
The default result of `st_intersects` is a *sparse matrix* - actually a kind of list. For each point (in the first argument of the function), it provides the row number of the second geometry argument where the geometries intersect. In our case, where the points lie inside the circle (also, in this case, `circle3` is a single row). So, for the first point of `rnd_pts`, the result is 1 as it intersects (or lies within) the first (and only) polygon in `circle3`. The second two points (row 2 and 3) are empty as they do not intersect any polygons. To get the points that are inside the circle, we can ask which of the results are not empty.  

The `lengths()` function gives us the length (or number of items) of each elemennt of a list. Here, this will either be 1, where a point intersects the circle, or it will be empty (where it doesn't). 
```{r clip4, exercise =  TRUE, exercise.setup = "clip3"}

lengths(pts_intersect)

```

If we ask which of the lengths are greater than 0, we will have the row number (or index) of each intersecting point (in the circle).

```{r clip5, exercise =  TRUE, exercise.setup = "clip3"}

lengths(pts_intersect) > 0

```

We can now use this results to subset (or index) our rnd_pts (using the same `[,]` indexing as previously). For example, we could just plot those points inside the circle:

```{r clip6, exercise =  TRUE, exercise.setup = "clip3"}

plot(st_geometry(circle3))
plot(st_geometry(rnd_pts[lengths(pts_intersect) > 0,]), add = T)

```

Alternatively, we could colour the points seperately with those inside or outside the circle. We could either do this directly with the `col` argument of the `plot)()` function, or we could add a new column to the `rnd_pts` object to signal which points are inside the circle.

```{r clip7, exercise =  TRUE, exercise.setup = "clip3"}

plot(st_geometry(rnd_pts), 
     col = (lengths(pts_intersect) > 0) + 1, 
     pch = 16,
     ylim = c(0,15))

plot(st_geometry(circle3), add = T)

```

