---
title: "Comparing between groups"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```


# Comparing between groups  

## Example workflow

*Here's an example analysis where we want to compare between two groups.*

Run the R code to follow the story.

You are studying the rare, threatened Tee bird in South America. As part of some basic ecological analysis you want to know whether there is a difference in weight between sexes as this might inform different conservation strategies. You have a data set of some female and male weights from captured birds.

We are going to use the analysis protocol to work through this example.

-*1 Ask your question*
Is there a difference in weight between female and male Tee birds?
(i.e. comparing mean weight between two groups)

-2 *Check and explore your data*

-2.1 First import the data. 

```{r example-t-test-import, exercise=TRUE}

tBird <- read.csv("data/bird_captures.csv", stringsAsFactors = TRUE)

```

-2.2 Now explore the data. Use the `head()` and `str()` functions to check the data.
How many columns are there? What do they represent?

```{r example-t-test-explore, exercise=TRUE, exercise.setup = "example-t-test-import"}


```

```{r example-t-test-explore-solution, exercise.reveal_solution = TRUE}
head(tBird)
str(tBird) # Numeric and categorical (factor) data types

```

-2.3 Now plot the data. First plot just the weight column and check for any outliers.

```{r example-t-test-plot1, exercise=TRUE, exercise.setup = "example-t-test-import"}
# Index plots - any obvious outliers?
plot(tBird$weight)

```

A box plot is suitable for a numeric variable and a categorical (grouping) variable.
Write the code to plot the weight for each sex as a boxplot. 

```{r example-t-test-plot2, exercise=TRUE, exercise.setup = "example-t-test-import"}


```

```{r example-t-test-plot2-solution, exercise.reveal_solution = TRUE}

boxplot(weight ~ sex, data = tBird)

```


-3 *Write your Null Hypothesis - H0*

There is no difference between the mean weight of female and male Tee birds from a sample of 30 captures.

-4 *Decide on your stats test*

We have two groups (two levels of the sex factor) and a numeric (response) variable. To find whether there is a significant difference between groups could be assessed with a T-test or a Wilcoxon Test.

-4.1 Is our data normal? Check distribution visually. Notice how we are using the `tapply()` function do plot one histogram for each sex.
```{r example-t-test-decide-hist, exercise=TRUE, exercise.setup = "example-t-test-import"}

tapply(tBird$weight, tBird$sex, hist)

```

This can also be tested formally, but let's go by the histograms and assume they are normally distributed.

-5 *Do test and check significance value*
```{r example-t-test-doTest, exercise=TRUE, exercise.setup = "example-t-test-import"}
# Given that our data are normally distributed, we can use the t-test:
t.test(weight ~ sex, data = tBird) # note this is the same format as the boxplot

```


-6 *Present your results*

-6.1 Extract statistics, effect size, sample size.
```{r example-t-test-stats, exercise=TRUE, exercise.setup = "example-t-test-import"}
# For simple tests like this you can copy directly from the console (it's also possible to export as a csv)
# i.e. Copy the p-value, t statistic and the degrees of freedom (an indication of sample size)
t.test(weight ~ sex, data = tBird)
```

We also want to report the means, and standard deviations for each group. Write the code, using `tapply()` to do that.
```{r example-t-test-means, exercise=TRUE, exercise.setup = "example-t-test-import"}


```


```{r example-t-test-means-solution, exercise.reveal_solution = TRUE}
# We also want to report the means, and standard deviations for each group
tapply(tBird$weight, tBird$sex, mean)
tapply(tBird$weight, tBird$sex, sd)

```

-6.2  Make a plot the data for your report

Do a plot with all the correct axis labels (including units), axis limits, etc.
```{r example-t-test-plot, exercise=TRUE, exercise.setup = "example-t-test-import"}

boxplot(weight ~ sex, data = tBird,
        xlab = "Weight (g)",
        ylab = "Sex",
        main = "")

```

6.3 Write a results statement:

Using a t-test we found that, on average, female Tee birds were significantly heavier than males (t = 3.71, df = 57.9, p < 0.001) from a sample of 30 captures. Females had a higher mean weight (8.14g `r knitr::asis_output("\U00B1")`0.70) than males (7.45g `r knitr::asis_output("\U00B1")` 0.74).

## Comparing more than two groups






```{r quiz1}
quiz(
  question_numeric("How many rows does the trees data set have?",
                   answer(31, correct = TRUE)
  ),
  question("What are the names of the columns of the trees data set?",
    answer("Girth", correct = TRUE),
    answer("height"),
    answer("Volume", correct = TRUE),
    answer("Mass"),
    answer("Height", correct= TRUE)
  ),
  question_radio("What is the height of the third tree in the data set?",
                 answer(65),
                 answer(63, correct = TRUE),
                 answer(31),
                 answer(10.2),
                 answer(8.8),
                 random_answer_order = TRUE,
                 allow_retry = TRUE)
  
)
```
